Schema Evolution in Interactive Programming Systems

### 2024-10-09 paper.tex
Update figures, final version layout tweaks

\documentclass[english,submission]{programming}

@@

\author{Anonymous}
\affiliation{blinded}
% \author[a]{Jonathan Edwards}%[0000-0003-1958-7967]
% \authorinfo{is an independent researcher. Contact him at \email{jonathanmedwards@gmail.com}.}
% \affiliation[a]{Independent, Boston, MA, USA}

% \author[b]{Tomas Petricek}%[0000-0002-7242-2208]
% \authorinfo{is an Associate Professor at Charles University. Contact him at \email{tomas@tomasp.net}.}
% \affiliation[b]{Charles University, Prague, Czechia}

% \author[c,d]{Tijs van der Storm}%[0000-0001-8853-7934]
% \authorinfo{is a Senior Researcher at Centrum Wiskunde \&\ Informatica (CWI) and a Professor at the University of Groningen. Contact him at \email{storm@cwi.nl}.}
% \affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
% \affiliation[d]{University of Groningen, Groningen, Netherlands}

% \author[e]{Geoffrey Litt}%[0000-0003-0858-5165]
% \authorinfo{is a Senior Researcher at the independent research lab Ink \& Switch. Contact him at \email{gklitt@gmail.com}.}
% \affiliation[e]{Ink \& Switch, Planet Earth}

% \authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}

@@

% TODO: Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification System

@@

Many improvements to programming have come from shortening feedback loops, for example with Integrated Development Environments, Unit Testing, Live Programming, and Distributed Version Control.
A barrier to feedback that deserves greater attention is \emph{Schema Evolution}.
When requirements on the shape of data change then existing data must be migrated into the new shape, and existing code must be modified to suit. Currently these adaptations are often performed manually, or with ad hoc code. Manual schema evolution not only delays feedback but since it occurs outside the purvey of version control tools it also interrupts collaboration.

@@

Many improvements to the practice of programming have come from speeding up feedback loops. Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming. Unit tests and interactive type checking provide faster feedback on errors that modern programmers now expect. Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing. Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
However there is still much room to improve feedback loops throughout software development.

In this paper we identify a common problem, \emph{Schema Change}, that interferes with feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations ~\cite{erhard06}. We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice, which we discuss further in \S\ref{sec:related} Related Work. While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts. We hope these challenge problems serve as examples to ground future research and as a basis for evaluating its results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

\begin{wrapfigure}{r}{12em}
\vspace{-1.5em}
\includegraphics[width=12em]{figures/arr-basic.png}
\vspace{-1.5em}
\end{wrapfigure}

The situation is illustrated on the right. We start with code $C_1$ and
data $D_1$ that have a shape corresponding to schema $T_1$. When the schema changes from $T_1$ to $T_2$ ($\rightarrow$),
a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
of \emph{schema evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
to the new schema $T_2$.
%
The complexity of the problem varies and finding a suitable transformation may require user
input. If there is code, it typically needs to be transformed, but data can sometimes be
discarded. We refer to such transformations done by a single user as \emph{local} and draw
them inside a box.

\begin{wrapfigure}{r}{12em}
\vspace{-1.5em}
\includegraphics[width=12em]{figures/arr-forkjoin.png}
\vspace{-1.5em}
\end{wrapfigure}

Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
variants that exist independently and changes made to them may need to be synchronized. The
situation is illustrated on the right. Here, the initial schema $T_1$ is transformed by two
users independently into schema $T_2$ and $T_3$. The corresponding data is transformed accordingly,
producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
obtaining a schema $T_4$ and data $D_4$ that adopt both of the changes.

@@

\paragraph{Locality of Change.}
We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
The \emph{local dimension} is concerned with how schema evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how schema evolution
propagates across different program variants that may exist independently.

\paragraph{Program Layers.}
We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
data that are structured by those schema. In specific case studies, the layers may map to
different aspects of programs, some may be not present, and some may be implicit.

\begin{itemize}
\item \emph{Schema} -- structures program data and determines aspects of code. A change in schema
  requires a corresponding change in data and code. Schema may be explicit (database schema, type declarations)
  or implicit (expected shape of data structures or a document).
\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
  database) or less permanent (state in live programming). When schemas evolve, data needs to be
  updated or (in case of transient state) discarded.
\item \emph{Code} -- or program logic, but not including source code that defines schema.
  When schemas evolve, code needs to change to consume/produce data of the correct shape.
\end{itemize}

\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their schema, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
schema evolution in future programming systems. We consider two characteristics.

@@

\paragraph{Challenge Problems.}
These dimensions, layers, and variants generate a wide range of configurations for challenge problems, but
our aim is not to cover all of them. Instead we use this perspective to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain, and we hope they provide an inspiration and a benchmark for
developers of future programming systems.

@@

\noindent
There is a number of edits to the types that the programmer may now want to do without
restarting the application and losing the data. For example, they may want to change the
\texttt{completed} field of \texttt{Item} to instead store optional completion time:

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; completed : Maybe<DateTime> }
type State = { items : Item list }
type Event =
  | .. | SetCompleted of id * Maybe<DateTime> | ..
\end{lstlisting}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0em}
\includegraphics[width=12em]{figures/arr-steps.png}
\vspace{0em}
\end{wrapfigure}
\frameworkboxtitle{Primitive Schema Transformations}
In this case study, the evolving schema is the \texttt{State} type. When the type evolves,
the program logic (code) and the current live value of the type (data) needs to evolve
correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
primitive code and data transformation. The primitive transformations may either be invoked
through a user interface, or inferred from textual code edits made by the programmer.
}

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), such past work
does not resolve all problems in our second challenge.

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{-0.5em}
\includegraphics[width=12em]{figures/arr-db.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Equivalence in Schema Evolution}
%We described our conceptual framework in terms of types, code and data, but this case study shows that the same patterns appear elsewhere.
In database systems, the schema
$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
in work on schema evolution. Extract/Absorb Entity operations are also interesting because
they transform an equivalent pair of schemas and so it should be possible to apply them
(and the corresponding data and query transformations) automatically.
}

@@

The challenge focuses on collaborative conference planning, illustrated in Figure~\ref{fig:conf}
and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
a conference that will feature talks by several invited speakers. They need to agree on the
speaker list, contact speakers and calculate the conference budget.
%
The ideas in this section also apply to work on computational scientific notebooks
that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoing
work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
and Living Papers~\cite{Heer2023}.

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-0.5em}
\includegraphics[width=13em]{figures/arr-datafork.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Merging Data and Code Edits}
In this example, the \emph{schema} is implicit and represents the structure of the document.
It involves two independent edits. In one (top right), the data $D_1$ is
changed to $D_2'$, but the schema $T_1$ remains unchanged. In the other (bottom left), the
schema $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
of schema $T_2$. In the next section, we will see the same pattern, but also involving
code edits. We assume the \emph{convergence} model in this section, but will discuss
the more challenging \emph{divergence} model later in the paper.}

@@

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new schema to an older schema.
In the diagram on the right, we only show data and schema, ignoring the code added by the
accounting department. We start with data $D_1$ of schema~$T_1$. The accounting department
evolves the schema (via Extract Entity), obtaining data $D_2$ of schema $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtaining $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from schema $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
schema evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14.5em}
\vspace{-0.5em}
\includegraphics[width=14.5em]{figures/arr-divergence.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Maintaining Divergent Variants}
The extra complexity in the above scenario arises from the fact that both schema and
code are forked and evolve independently. In one branch, original schema $T_1$ evolve
into $T_2$ and original code $C_1$ into $C_3$. In another branch, original schema $T_1$
evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
the independently done changes into a new version of schema and code, written as
$T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
conflicting changes that would need to be resolved with assistance from the user.}

@@

Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the
write alone, it is impossible to make a perfect choice among these options. The programming system
thus needs to consult the user.

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-1em}
\includegraphics[width=13em]{figures/arr-backport.png}
\vspace{-2em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration (Revisited)}
The diagram illustrating the scenario remains the same as in the case of divergence control
for Extract Entity. The initial data and schema $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
generated for the schema evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
In case of Extract Entity this was in principle possible.}

@@

Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this
problem would give developers or users tools to manage these tradeoffs. For example, a system
might allow developers to specify the desired behavior for a particular pair of schemas based
on the requirements of the domain. Or a system might even allow users to manually
disambiguate their intent for a given write.
Whatever the solution, it must be easy to learn and simple to use otherwise it won't be used in practice.

@@

@@

~

We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type AKA schema) that
defines the AST structure of the domain-specific language.

@@

@@

\begin{table}[t]
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add/Rename & state & Simple  \\
Remove & state & Structurally simple, but heuristic if state is current \\
Add & variable & Add entry to map and initialize \\
Remove & variable & Remove from the variable map (NB: assumes variable is unused)\\
Rename & variable & Rename in class, preserving value\\
Type change & variable & Migrate value in map \\
Add & transition & Simple \\
Remove & transition & Structurally simple, but heuristic for pending events\\
Event change & transition & Structurally simple, but heuristic for pending events\\
Add/remove & statement & Hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\end{adjustwidth}
\vspace{-1em}
\end{table}

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as schema, code and data co-evolution as in the case of the Elm
architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

When the state machine is instantiated, the metamodel and the definition are used to derive
the run-time type ${}^+\!T_0$. The run-time data ${}^+\!D_0$ conforming to this type still
contain the state machine definition, but also the necessary run-time state required for
its execution, which will produce new versions of the run-time data such as ${}^+\!D_0'$.

If the state machine definition changes from $D_0$ to $D_1$ (top right), we can synthesize
a new run-time state machine type ${}^+\!T_1$ and new data ${}^+\!D_1$ conforming to it.
The challenge is to merge the two run-time types instantiated from different definitions
to produce a type ${}^+\!T_0 \diamond {}+\!T_1$, but also merge the run-time state of the
running machine with the new instantiated state to obtain ${}^+\!D'_0 \diamond {}^+\!D_1$.
}

@@

\emph{Local-first software}~\cite{localfirst} adds the divergence dimension to NoSQL evolution, because the code running on different replicas can evolve at different paces. \citet{Cambria} applies lenses~\cite{Foster2007} to this problem, specifically discussing the \emph{Multiplicity Change} problem.

Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}. Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}. Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.

>>>>

\documentclass[english,crc]{programming}

@@

\author[a]{Jonathan Edwards}[0000-0003-1958-7967]
\authorinfo{is an independent researcher working on drastically simplifying programming. He is known
for his Subtext series of programming language experiments and his blog at \href{https://alarmingdevelopment.org}{\texttt{alarmingdevelopment.org}}.
He has been a researcher at MIT CSAIL and CDG/HARC. He tweets \href{https://x.com/jonathoda}{\texttt{@jonathoda}} and can be reached at
\email{jonathanmedwards@gmail.com}.}
\affiliation[a]{Independent, Boston, MA, USA}

\author[b]{Tomas Petricek}[0000-0002-7242-2208]
\authorinfo{is an assistant professor at Charles University. He
is interested in finding easier and more accessible ways of thinking
about programming. To do so, he combines technical work on
programming systems and tools with research into history and
philosophy of science. His work can be found at \href{https://tomasp.net}{\texttt{tomasp.net}} and
he can be reached at \email{tomas@tomasp.net}.}
\affiliation[b]{Charles University, Prague, Czechia}

\author[c,d]{Tijs van der Storm}[0000-0001-8853-7934]
\authorinfo{is a senior researcher in the Software Analysis and Transformation (SWAT) group at Centrum Wiskunde \&\ Informatica (CWI) in Amsterdam,
and full professor in Software Engineering at the University of Groningen in Groningen. His research focuses on improving programmer
experience through new and better software languages and developing the tools and techniques to engineer them in a modular and interactive fashion. Contact him at \email{storm@cwi.nl}.}
\affiliation[c]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, Netherlands}
\affiliation[d]{University of Groningen, Groningen, Netherlands}

\author[e]{Geoffrey Litt}[0000-0003-0858-5165]
\authorinfo{is a senior researcher at the independent research lab Ink \& Switch.
Previously, he completed a PhD at MIT CSAIL in the Software Design Group advised by Daniel Jackson.
Before that, he spent five years as an early engineer and designer building the edtech startup Panorama Education (YC S13).
Contact him at \email{gklitt@gmail.com}.}
\affiliation[e]{Ink \& Switch, Planet Earth}

\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}

@@

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011008</concept_id>
       <concept_desc>Software and its engineering~General programming languages</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
       <concept_desc>Software and its engineering~Data types and structures</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011073</concept_id>
       <concept_desc>Software and its engineering~Software maintenance tools</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10002951.10002952.10003219</concept_id>
       <concept_desc>Information systems~Information integration</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[300]{Software and its engineering~General programming languages}
\ccsdesc[500]{Software and its engineering~Data types and structures}
\ccsdesc[300]{Software and its engineering~Software maintenance tools}
\ccsdesc[300]{Information systems~Information integration}

@@

Many improvements to programming have come from shortening feedback loops, for example with Integrated Development Environments, Unit Testing, Live Programming, and Distributed Version Control.
A barrier to feedback that deserves greater attention is \emph{Schema Evolution}.
When requirements on the shape of data change then existing data must be migrated into the new shape, and existing code must be modified to suit. Currently these adaptations are often performed manually, or with ad hoc scripts. Manual schema evolution not only delays feedback but since it occurs outside the purview of version control tools it also interrupts collaboration.

@@

Many improvements to the practice of programming have come from speeding up feedback loops. Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming. Unit tests and interactive type checking provide faster feedback on errors that modern programmers now expect. Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing. Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit} to convey changes between coworkers with less friction.
However there is still much room to improve feedback loops throughout software development.

In this paper we identify a common problem, \emph{Schema Change}, that interferes with feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necessitates manual and error-prone data migrations ~\cite{erhard06}. We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in Section~\ref{sec:related}. In general, when the shape of data becomes a dependency, changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to \emph{Interactive Programming Systems}~\cite{techdims} and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice. While some of them come from our individual research efforts we have attempted to make neutral problem statements independent of our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed, the aim of this paper is to provide a basis for collaboration between such related efforts, facilitate further research, as well as to serve as a basis for evaluating research results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

\begin{wrapfigure}{r}{12em}
\vspace{-1.6em}
\includegraphics[width=12em]{figures/arr-basic.png}
\vspace{-1.6em}
\end{wrapfigure}

The situation is illustrated on the right. We start with code $C_1$ and
data $D_1$ that have a shape corresponding to schema $S_1$. When the schema changes from $S_1$ to $S_2$ ($\rightarrow$),
a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
of \emph{schema evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
to the new schema $S_2$.
%
The complexity of the problem varies and finding a suitable transformation may require user
input. If there is code, it typically needs to be transformed, but data can sometimes be
discarded. We refer to such transformations done by a single user as \emph{local} and draw
them inside a box.

\begin{wrapfigure}{r}{12em}
\vspace{-1.5em}
\includegraphics[width=12em]{figures/arr-forkjoin.png}
\vspace{-2em}
\end{wrapfigure}

Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
variants that exist independently and changes made to them may need to be synchronized. The
situation is illustrated on the right. Here, the initial schema $S_1$ is transformed by two
users independently into schema $S_2$ and $S_3$. The corresponding data is transformed accordingly,
producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
obtaining a schema $S_4$ and data $D_4$ that adopt both of the changes.

@@

\vspace{-0.4em}
\paragraph{Locality of Change.}
We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
The \emph{local dimension} is concerned with how schema evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how schema evolution
propagates across different program variants that may exist independently.

\vspace{-0.4em}
\paragraph{Program Layers.}
We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
data that are structured by those schema. In specific case studies, the layers may map to
different aspects of programs, some may be not present, and some may be implicit.

\begin{itemize}
\item \emph{Schema} -- structures program data and determines aspects of code. A change in schema
  requires a corresponding change in data and code. Schema may be explicit (database schema, type declarations)
  or implicit (expected shape of data structures or a document).
\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
  database) or less permanent (state in live programming). When schemas evolve, data needs to be
  updated or (in case of transient state) discarded.
\item \emph{Code} -- or program logic, but not including source code that defines schema.
  When schemas evolve, code needs to change to work with data of the correct shape.
\end{itemize}

\vspace{-0.4em}
\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their schema, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
schema evolution in future programming systems. We consider two characteristics.

@@

\vspace{-0.4em}
\paragraph{Challenge Problems.}
These dimensions, layers, and variants generate a wide range of problem configuration.
Our aim is not to cover all of them. Instead we use this perspective to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain. We hope they provide an inspiration and a benchmark for
developers of future systems.

@@

There is a number of edits to the types that the programmer may now want to do without
restarting the application and losing the data. For example, they may want to change the
\texttt{completed} field of \texttt{Item} to instead store optional completion time:

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; completed : Maybe<DateTime> }
type State = { items : Item list }
type Event = .. | SetCompleted of id * Maybe<DateTime> | ..
\end{lstlisting}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0em}
\includegraphics[width=12em]{figures/arr-steps.png}
\vspace{0em}
\end{wrapfigure}
\frameworkboxtitle{Primitive Schema Transformations}
Here, the evolving schema is the \texttt{State} type. When the type evolves,
the program logic (code) and the current live value (data) needs to evolve
correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
primitive code and data transformation. The primitive transformations may either be invoked
through a UI, or inferred from textual code edits made by the programmer.
}

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (Section~\ref{sec:related}), such past work
does not resolve all problems in our second challenge.

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{-2em}
\includegraphics[width=12em]{figures/arr-db.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Equivalence in Schema Evolution}
%We described our conceptual framework in terms of types, code and data, but this case study shows that the same patterns appear elsewhere.
In database systems, the schema
$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
in work on schema evolution. Extract/Absorb Entity are also interesting because
they transform an equivalent pair of schemas and so it should be possible to apply them
(and the corresponding data and query transformations) automatically.
}

@@

The challenge focuses on collaborative conference planning, illustrated in Figure~\ref{fig:conf}
and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
a conference that will feature talks by several invited speakers. They need to agree on the
speaker list, contact speakers and calculate the conference budget.
%
The ideas in this section also apply to computational scientific notebooks
that combine code and data \cite{Kluyver2016}. We hope to contribute to ongoing
work advancing the format \cite{Nextjournal21,Crichton2021,Heer2023}.

@@

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-0.5em}
\includegraphics[width=13em]{figures/arr-datafork.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Merging Data and Code Edits}
In this example, the \emph{schema} is implicit and represents the structure of the document.
It involves two independent edits. In one (top right), the data $D_1$ is
changed to $D_2'$, but the schema $S_1$ remains unchanged. In the other (bottom left), the
schema $S_1$ and data $D_1$ co-evolve into $S_2$ and $D_2$ and the data is further changed
to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
of schema $S_2$. In the next section, we will see the same pattern, but also involving
code edits. We assume the \emph{convergence} model in this section, but will discuss
the more challenging \emph{divergence} model later in the paper.}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new schema to an older schema.
In the diagram on the right, we only show data and schema, ignoring the code added by the
accounting department. We start with data $D_1$ of schema~$S_1$. The accounting department
evolves the schema (via Extract Entity), obtaining data $D_2$ of schema $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtaining $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from schema $S_2$ to
$S_1$ using an automatically generated backward transformation. That is,
schema evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14.5em}
\vspace{-0.5em}
\includegraphics[width=14.5em]{figures/arr-divergence.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Maintaining Divergent Variants}
The extra complexity in the above scenario arises from the fact that both schema and
code are forked and evolve independently. In one branch, original schema $S_1$ evolve
into $S_2$ and original code $C_1$ into $C_3$. In another branch, original schema $S_1$
evolve into $S'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
the independently done changes into a new version of schema and code, written as
$S_2 \diamond S'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
conflicting changes that would need to be resolved with assistance from the user.}

@@

Because the intent of the writer to the scalar schema cannot be unambiguously interpreted from the
write alone, it is impossible to make a perfect choice among these options. A programming system could consult the user to disambiguate their intent, as well as choose (and explain) a default behavior.

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-1em}
\includegraphics[width=13em]{figures/arr-backport.png}
\vspace{-2em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration (Revisited)}
The diagram illustrating the scenario remains the same as in the case of divergence control
for Extract Entity. The initial data and schema $D_1, S_1$ co-evolve into $D_2, S_2$ (adding
multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
generated for the schema evolution (from $S_1$ to $S_2$) cannot be generated fully automatically.
In case of Extract Entity this was in principle possible.}

@@

Although there is likely no silver bullet to navigate these tradeoffs, a good solution
would give developers or users tools to manage these tradeoffs. For example, a system
might allow developers to specify the desired behavior for a particular pair of schemas based
on the requirements of the domain. Or a system might even allow users to manually
disambiguate their intent for a given write.
Whatever the solution, it must be easy to learn and simple to use otherwise it won't be used in practice.

@@

We considered live programming in the context of Elm in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type, or, a schema) that
defines the AST structure of the domain-specific language.

@@

@@

\begin{table}[b]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add/Rename & state & Simple  \\
Remove & state & Structurally simple, but heuristic if state is current \\
Add & variable & Add entry to map and initialize \\
Remove & variable & Remove from the variable map (assuming it is unused)\\
Rename & variable & Rename in class, preserving value\\
Type change & variable & Migrate value in map \\
Add & transition & Simple \\
Remove & transition & Structurally simple, but heuristic for pending events\\
Event change & transition & Structurally simple, but heuristic for pending events\\
Add/remove & statement & Hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\vspace{-1em}
\end{table}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as schema, code and data co-evolution as in the case of the Elm
architecture in Section~\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $S$ defined by the metamodel.

When the state machine is instantiated, the metamodel and the definition are used to derive
the run-time type ${}^+\!S_0$. The run-time data ${}^+\!D_0$ conforming to this type still
contain the state machine definition, but also the necessary run-time state required for
its execution, which will produce new versions of the run-time data such as ${}^+\!D_0'$.

If the state machine definition changes from $D_0$ to $D_1$ (top right), we can synthesize
a new run-time state machine type ${}^+\!S_1$ and new data ${}^+\!D_1$ conforming to it.
The challenge is to merge the two run-time types instantiated from different definitions
to produce a type ${}^+\!S_0 \diamond {}+\!S_1$, but also merge the run-time state of the
running machine with the new instantiated state to obtain ${}^+\!D'_0 \diamond {}^+\!D_1$.
}

@@

\emph{Local-first software}~\cite{localfirst} adds the divergence dimension to NoSQL evolution, because the code running on different replicas can evolve at different paces. \citet{Cambria} applies lenses~\cite{Foster2007} to this problem, specifically discussing the \emph{Multiplicity Change} problem.
%
Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}. Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}. Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.
